---
title: "Class 8 Mini-Project"
author: "Mari Williams (A15858833)"
format: gfm
toc: TRUE
---

## Data prep


```{r}
fna.data <- "WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names=1)
head(wisc.df)
```
```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df$diagnosis
diagnosis <-as.factor(diagnosis)
```
***Q1. How many observations are in this dataset?***

```{r}
dim(wisc.data)
```
569 observations

***Q2. How many of the observations have a malignant diagnosis?***
```{r}
sum(diagnosis=="M")
```
212 observations are diagnosed as malignant

***Q3. How many variables/features in the data are suffixed with _mean***
```{r}
length(grep("_mean", colnames(wisc.data)))

```
10 variables are suffixed with _mean


## Principal Component Analysis
```{r}
colMeans(wisc.data)

apply(wisc.data,2,sd)
```

```{r}
wisc.pr <- prcomp(wisc.data, center = TRUE, scale. = TRUE)
summary(wisc.pr)
```
***Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?*** 

44% of the variance is explained by PC1

***Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?***

3 PCs

***Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?***

7 PCs

## Interpreting PCA results
```{r}
biplot(wisc.pr)
```


***Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?***
Text and numbers are indecipherable



```{r}
library(ggplot2)
ggplot(wisc.pr$x) + aes(PC1, PC2, col = diagnosis, xlab = "PC1", ylab= "PC2") + geom_point()
```
***Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?***

```{r}
ggplot(wisc.pr$x) + aes(PC1, PC3, col = diagnosis, xlab = "PC1", ylab= "PC3") + geom_point()
```
Both plots have pretty good separation between diagnoses but the first one is a little cleaner. Similar x distributions, the y distributions are shifted. 

## Variance explained

```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```
```{r}
pve <- pr.var / sum(pr.var)
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

### Communicating PCA Results
***Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC.***

```{r}
lv <- wisc.pr$rotation[,1]
lv["concave.points_mean"]
```
-0.261


## Hierarchical clustering
```{r}
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist, method= "complete")
```
***Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?***

```{r}
plot(wisc.hclust)
abline(a=19, b=0, col="red", lty=2)
```
height=19

***Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.***
```{r}
clust <- function(data, way) {
  wisc.hclust <- hclust(data, method= way)
  plot(wisc.hclust, xlab=way)

}
sapply(c("complete", "single", "average", "ward.D2"), function(m) clust(data.dist, m))



```
I like ward.D2 as well, because of the clean separation of the clusters


## Combining methods
```{r}
d<- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:7]), method= "ward.D2")
plot(wisc.pr.hclust)
```

```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:3]), method= "ward.D2")
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
table(grps, diagnosis)

```
TP: 179
FP:33

```{r}
g <- as.factor(grps)
levels(g)
g <- relevel(g,2)
levels(g)
plot(wisc.pr$x[,1:2], col=g)
plot(wisc.pr$x[,1:2], col=diagnosis)


```

```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method="ward.D2")

wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
table(wisc.pr.hclust.clusters)
```
***Q13. How well does the newly created model with four clusters separate out the two diagnoses?***
```{r}
table(wisc.pr.hclust.clusters, diagnosis)
```
Pretty well! 9% of the assignments do not agree, but the rest do.


 ***Q14. How well do the hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.***

 
```{r}
clust <- function(data, way) {
wisc.hclust <- hclust(data, method= way)
wisc.pr.hclust.clusters <- cutree(wisc.hclust, k=2)

tbl<- (table(wisc.pr.hclust.clusters, diagnosis))
plt<- (plot(wisc.pr$x[,1:2], col=wisc.pr.hclust.clusters, xlab=way))

return(list(table = tbl, plot = plt))

}

result <- sapply(c("complete", "single", "average", "ward.D2"), function(m) clust(data.dist, m))
result[1,1:4]
result$plot




```
Most of the methods except ward.D2 have very innacurate categorization that does not overlap with the diagnosis. 