---
title: "Halloween project"
author: "Mari Williams (A15858833)"
format: gfm
---

```{r}
candy <- read.csv(url("https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"), row.names=1)
head(candy)
```
***Q1. How many different candy types are in this dataset?***
```{r}
nrow(candy)
ncol(candy)
```
`r nrow(candy)` different candies categorized in `r ncol(candy)` descriptors

***Q2. How many fruity candy types are in the dataset?*** 
```{r}
sum(candy$fruity)
```
`r sum(candy$fruity)`


***Q3. What is your favorite candy in the dataset and what is it’s winpercent value?***
```{r}

candy["Caramel Apple Pops", ]$winpercent

```
Caramel Apple Pops winpercent is `r candy["Caramel Apple Pops", ]$winpercent`


***Q4. What is the winpercent value for “Kit Kat”?***
```{r}
candy["Kit Kat", ]$winpercent
```

***Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?***
```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```
```{r}
library("skimr")
#skim(candy)
```
The skimr package was giving me issues with rendering; I ran it but had to take it out for rendering

***Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?***

The last column, winpercent, is scaled based off of 100% versus a 0 or 1 value

***Q7. What do you think a zero and one represent for the candy$chocolate column?***

0 means the candy is not classified as chocolate, 1 means it is


***Q8. Plot a histogram of winpercent values***

```{r}
hist(candy$winpercent, 30)
```

***Q9. Is the distribution of winpercent values symmetrical?***

Skewed more towards the lower end than truly symmetrical. 

***Q10. Is the center of the distribution above or below 50%?***

Below 50%

***Q11. On average is chocolate candy higher or lower ranked than fruit candy?***

```{r}
mean(candy$winpercent[as.logical(candy$fruity)])
mean(candy$winpercent[as.logical(candy$chocolate)])
```
Chocolate is on average ranked higher than fruity candy. 


***Q12. Is this difference statistically significant?***

```{r}
t.test(candy$winpercent[as.logical(candy$fruity)], candy$winpercent[as.logical(candy$chocolate)])
```

Yes, it is statistically significant

```{r}
library(dplyr)

```

***Q13. What are the five least liked candy types in this set?***
```{r}
rownames(candy %>% 
  arrange(winpercent) %>%
  head(5))
```

***Q14. What are the top 5 all time favorite candy types out of this set?***
```{r}
rownames(candy %>% 
  arrange(-winpercent) %>%
  head(5))
```

***Q15. Make a first barplot of candy ranking based on winpercent values.***

```{r fig.width = 10, fig.height = 15}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col(width = .8)
```

***Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?***


```{r fig.width = 10, fig.height = 15}

ggplot(candy) + 
 aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(width = .8)
```

```{r fig.width = 10, fig.height = 15}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```
***Q17. What is the worst ranked chocolate candy?***

Sixlets

***Q18. What is the best ranked fruity candy?***

Starburst

### Price Percent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```


***Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?***
```{r}
ord <- order(candy$winpercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
Reese's Mini's are the cheapest among the most popular (But I would argue less bang for your buck overall, since its a miniature size of Reese)


Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )

```
Nik L Nip is the least liked out of the most expensive. 

### Correlation

```{r}
library(corrplot)

```
```{r}
cij <- cor(candy)
corrplot(cij)
```
***Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?***

Chocolate and fruity are anti-correlated, as well as bar and fruity, and bar and pluribus. (I like chocolate fruit candies, like chocolate covered cherries!)

***Q23. Similarly, what two variables are most positively correlated?***

Bar and chocolate or chocolate and winpercent

### PCA

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
pc3 <- cbind(candy, pca$x[,1:3])
p <- ggplot(pc3) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(pc3),
            label=rownames(pc3)) +
        geom_point(col=my_cols)

p
```
```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

***Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?***

Fruity and pluribus. I thought chocolate and bar would be picked up more, but the fruity candies cluster strongly on the plot so this makes sense. 