---
title: "Pertussis Mini Project"
author: "Mari Williams (A15858833)"
format: gfm
---

```{r, include=FALSE}
cdc <- data.frame(
                          Year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,
                                   2019L,2020L,2021L,2022L,2023L),
  No..Reported.Pertussis.Cases = c(107473,
                                   164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617,6124,
                                   2116,3044,7063)
)
cdc
```
###Q1

```{r}
library(ggplot2)
ggplot(cdc) +
  geom_line(aes(x = Year, y = No..Reported.Pertussis.Cases)) +
  labs(title = "Reported Pertussis Cases in the US (1922-2023)",
       x = "Year",
       y = "Number of Reported Cases") +
  theme_minimal()
```
###Q2
```{r}
ggplot(cdc) +
  geom_line(aes(x = Year, y = No..Reported.Pertussis.Cases)) +
  labs(title = "Reported Pertussis Cases in the US (1922-2023)",
       x = "Year",
       y = "Number of Reported Cases") +
  theme_minimal() + geom_vline(xintercept = 1946, linetype="dashed", 
              color = "red", size=.5) +
  annotate("text", x = 1950, y = 250000, label = " wP ", color = "red") + geom_vline(xintercept = 1996, linetype="dashed", 
              color = "blue", size=.5) +
  annotate("text", x = 2000, y = 250000, label = "aP ", color = "blue") +geom_vline(xintercept = 2020, linetype="dashed", 
              color = "gray", size=.5) +
  annotate("text", x = 2020, y = 250000, label = " COVID ", color = "gray") 

```

###Q3
There is gradual increase in pertussis cases after the aP introduction, followed by a few spikes. This is likely due to a growing anti-vaccination movement in the US. It is also due to the fact that aP vaccines do not provide as long-lasting immunity as wP vaccines. The decrease in cases during the COVID-19 pandemic is likely due to social distancing and mask-wearing measures that were implemented to reduce the spread of COVID-19, which also reduced the spread of pertussis.

```{r}
(library(jsonlite))
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 

```
```{r}
head(subject, 3)

```
##Q4, Q5, Q6

```{r}
table(subject$infancy_vac)
table(subject$biological_sex)
table(subject$biological_sex, subject$race)

```

###Q7 and Q8

```{r}
library(lubridate)
for (i in 1:(length(subject$year_of_birth))) {
  subject$age_years[i] <- as.numeric(time_length( today() - ymd(subject$year_of_birth[i]),  "years"))
}

```
```{r}
library(dplyr)
ap <- subject %>% filter(infancy_vac == "aP")
mean(ap$age_years, na.rm = TRUE)

wp <- subject %>% filter(infancy_vac == "wP")
mean(wp$age_years, na.rm = TRUE)
```
```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```


```{r}
ggplot(subject) +
  aes(time_length(age_years, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```
Yes, wP group is older and more spread out


```{r}
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api/plasma_ab_titer", simplifyVector = TRUE) 
head(specimen)
```
##Q6

##Q7
```{r}
unique(titer$antigen)
```
###Q8
```{r}
ggplot(titer)+ aes(MFI, antigen) +
  geom_boxplot()
```



###Q9 and Q10

```{r}
library(dplyr)
meta <- inner_join(specimen, subject)
head(meta)
dim(meta)
abdata<- inner_join(meta, titer)
head(abdata)
dim(abdata)

```
###Q11

```{r}
table(abdata$isotype)
```

###Q12

```{r}
table(abdata$dataset)
```
There's 4 different dataset years. They decrease in size from 2020 to 2023, with 2023 being the smallest.

##Q13
```{r}
igg <- abdata %>% filter(isotype == "IgG")
ggplot(igg) +
  aes(x = MFI_normalised, y = antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

### Q14 Time course analysis
```{r}
abdata.tc <- abdata %>% filter(dataset == "2021_dataset")

abdata.tc %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2023 dataset IgG PT")
```
